---
title: "Comparing the Wilcoxon-Mann-Whitney to alternative statistic tests. "
output: 
  md_document:
    variant: markdown_phpextra
    preserve_yaml: TRUE
---
Supplementary Information for "Detect issue heterogenity in gene expression data with [*BioQC*](https://github.com/Accio/BioQC)" ([Jitao David Zhang](mailto:jitao_david.zhang@roche.com), Klas Hatje, Clemens Broger, Martin Ebeling and [Laura Badi](laura.badi@roche.com))


```{r setup, include=FALSE}
options(fig_caption=TRUE)
library(knitr)
opts_chunk$set(out.extra='style="display:block; margin: auto"', fig.align="center")
```

In this document, we show that the Wilcoxon-Mann-Whitney (WMW)-test is comparable or superior to alternative methods. 

*BioQC* is a R/Bioconductor package to detect tissue heterogeneity from high-throughput gene expression profiling data. It implements an efficient Wilcoxon-Mann-Whitney test, and offers tissue-specific gene signatures that are ready to use 'out of the box'.


```{r lib, warning=FALSE, message=FALSE, results="hide"}
library(testthat)
library(BioQC)
library(hgu133plus2.db) ## to simulate an microarray expression dataset
library(lattice)
library(latticeExtra)
library(gridExtra)
library(GEOquery)
library(xtable)
library(gplots)
library(reshape2)
library(ggplot2)
library(rbenchmark)

pdf.options(family="ArialMT", useDingbats=FALSE)

set.seed(1887)

## list human genes
humanGenes <- unique(na.omit(unlist(as.list(hgu133plus2SYMBOL))))

## read tissue-specific gene signatures
gmtFile <- system.file("extdata/exp.tissuemark.affy.roche.symbols.gmt",
                       package="BioQC")
gmt <- readGmt(gmtFile)
```

Two alternative methods could be compared with the Wilcoxon-Mann-Whitney test proposed by BioQC: the Kolmogonov-Smirnov test (KS-test hereafter), and the Student’s t-test  (t-test hereafter), or more particularly, the Welch’s test which does not assume equal sample number or equal variance, which is appropriate in the setting of gene expression studies. 

1. It is documented in statistics literature that the Wilcoxon-Mann-Whitney test offers a higher power than the Kolmogorov-Smirnov test. [^1][^2]. 
2. Compared with parameterized test methods such as the t-test, the Wilcoxon-Mann-Whitney test is (a) resistance to monotone transformation,  (b) suffers less from outliers, and (c) provides higher efficiency when many genes are profiled and the distribution of gene expression deviates from the normal distribution, which are important criteria in genome-wide expression data. 

Based on these considerations, BioQC implements a computationally efficient version of the Wilcoxon-Mann-Whitney test. In order not to confuse end-users, no alternative methods are implemented.

Nevertheless, in order to demonstrate the power of Wilcoxon-Mann-Whitney test in comparison with the KS-test or the t-test, we performed the sensitivity benchmark described in the [simulation studies](bioqc-simulation.html), for the two alternative tests respectively. 

```{r sensitivity_benchmark_fig, echo=FALSE, fig.width=8, fig.height=3, dev='svg', fig.cap="Sensitivity benchmark. Expression levels of genes in the ovary signature are dedicately sampled randomly from normal distributions with different mean values. The lines show the enrichment score for the Wilcoxon-Mann-Whitney test, the t-test and the Kolmogorov-Smirnov test respectively. In the right panel, noise was added by adding a value drawn from an additional normal distribution to a fraction of the genes."}
source("helper.R")

randomMatrixButOneSignature <- function(rows=humanGenes, signatureGenes,
                                        amplitudes=seq(0, 3, by=0.5)) {
  nrow <- length(rows)
  ncol <- length(amplitudes)
  mat <- matrix(rnorm(nrow*ncol),
                nrow=nrow, byrow=FALSE)
  rownames(mat) <- rows
  sigInd <- na.omit(match(signatureGenes, humanGenes))
  
  colClass <- factor(amplitudes)
  
  for(colInd in unique(colClass)) {
    isCurrCol <- colInd==colClass
    replaceMatrix <- matrix(rnorm(length(sigInd)*sum(isCurrCol),
                                  mean=amplitudes[isCurrCol][1]),
                            nrow=length(sigInd), byrow=FALSE)
    mat[sigInd, isCurrCol] <-  replaceMatrix
    }
  return(mat)
}

addNoise = function(matrix, fractionAffected=.1, stdv=1) {
  noise = matrix(rnorm(nrow(matrix)*ncol(matrix), sd=stdv), nrow=nrow(matrix), byrow=FALSE)
  addNoise = matrix(runif(nrow(matrix)*ncol(matrix)), nrow=nrow(matrix), byrow=FALSE) < fractionAffected
  matrix = matrix + addNoise*noise
  return(matrix)
}

do.test <- function(senseMat, tissueInd, test=t.test, alternative="greater") {
  bgInds = setdiff(1:nrow(senseMat), tissueInd)
  res = apply(senseMat, 2, function(col) {
    gs = col[tissueInd]
    bg = col[bgInds]
    return(test(gs, bg, alternative=alternative)$p.value)
  })
  return(res)
}


selGeneSet <- "Ovary_NGS_RNASEQATLAS_0.6_3"
selSignature <- gmt[[selGeneSet]]$genes
tissueInds <- sapply(gmt, function(x) match(x$genes, humanGenes))
senseAmplitudes <- rep(c(seq(0, 1, by=0.25), seq(1.5, 3, by=0.5)), each=50)
senseMat <- randomMatrixButOneSignature(rows=humanGenes,
                                        signatureGenes=selSignature,
                                        amplitudes=senseAmplitudes)
senseMatOutlier = addNoise(senseMat, stdv=15, fractionAffected=.01)

compare.tests = function(senseMat) {
  senseBioQC <- wmwTest(senseMat, tissueInds, valType="p.greater", simplify=TRUE)[selGeneSet,]
  senseTTest <- do.test(senseMat, tissueInds[[selGeneSet]], test=t.test)
  senseKS <- do.test(senseMat, tissueInds[[selGeneSet]], test=function(x, y, alternative) {
    ks.test(y, x, alternative=alternative)})
  
  comp = data.frame(BioQC=-log10(senseBioQC), t.test=-log10(senseTTest), ks.test=-log10(senseKS), senseAmplitudes=senseAmplitudes)
  comp.molten = melt(comp, id="senseAmplitudes")
  comp.summary = summarySE(comp.molten, measurevar="value", groupvars=c("senseAmplitudes", "variable"))
  return(comp.summary)
}

comp.summary = compare.tests(senseMat)
comp.summary.outlier = compare.tests(senseMatOutlier)

plot.comp = ggplot(comp.summary, aes(x=senseAmplitudes, y=value, color=variable)) + 
  geom_line() +
  geom_errorbar(aes(ymin=value-se, ymax=value+se), width=.1) + 
  geom_point() + xlab("Mean expression differnce") + ylab("Enrichment score") + ggtitle("without noise")

plot.comp.outlier = ggplot(comp.summary.outlier, aes(x=senseAmplitudes, y=value, color=variable)) + 
  geom_line() +
  geom_errorbar(aes(ymin=value-se, ymax=value+se), width=.1) + 
  geom_point() + xlab("Mean expression differnce") + ylab("Enrichment score") + ggtitle("with noise")

grid.arrange(plot.comp, plot.comp.outlier, ncol=2)
``` 


As expected, the results suggest, that both the KS-test and the WMW-test are robust to noise, while the performance of the t-test drops significantly on noisy data. Additionally, the WMW-test appears to be superior to the KS-test for low expression differences. 

References
----------
[^1]: Irizarry, Rafael A., et al. "Gene set enrichment analysis made simple."Statistical methods in medical research 18.6 (2009): 565-575.
[^2]: Filion, Guillaume J. "The signed Kolmogorov-Smirnov test: why it should not be used." GigaScience 4.1 (2015): 1.


R Session Info
----------------
```{r session_info}
sessionInfo()
```
