---
title: Compare Signatures derived from GTEx with BioQC
output:
  md_document:
    variant: markdown_phpextra
    preserve_yaml: TRUE
---

Here, we compare the overlap of the signatures derived from GTEx in an earlier step with those from BioQC derived with Gini index.


```{r libs, include=FALSE}
library(ribiosUtils)
libordie(BioQC)
libordie(gridExtra)
libordie(VennDiagram)
libordie(ggplot2)
libordie(reshape2)
libordie(RColorBrewer)
libordie(stringr)
libordie(MASS)
libordie(data.table)
options(fig_caption=TRUE)
library(knitr)
opts_chunk$set(out.extra='style="display:block; margin: auto"', fig.align="center") 
```

### Define functions and load data
```{r set_functions}
# functions for comparing sets. 
intersectSigs = function(gmt1, gmt2, sig1, sig2) {
  genes1 = gmt1[[sig1]]$genes
  genes2 = gmt2[[sig2]]$genes
  return(c(length(genes1), length(genes2), length(intersect(genes1, genes2))))
}

jaccInd = function(gmt1, gmt2, sig1, sig2) {
  is = intersectSigs(gmt1, gmt2, sig1, sig2)
  return(is[3]/(sum(is[1:2]) - is[3]))
}

drawVenn = function(gmt1, gmt2, sig1, sig2) {
  is = intersectSigs(gmt1, gmt2, sig1, sig2)
  plot.new()
  draw.pairwise.venn(is[1], is[2], is[3])
}

fisherTest = function(gmt1, gmt2, sig1, sig2) {
  is = intersectSigs(gmt1, gmt2, sig1, sig2)
  mat = matrix(c(is[3],is[2]-is[3],is[1]-is[3],sum(is[1:2])-is[3]), ncol=2)
  return(fisher.test(mat)$p.value)
}
```

```{r load_data}
# load signatures from BioQC
bioqc = readGmt(system.file("extdata/exp.tissuemark.affy.roche.symbols.gmt", package="BioQC"))
gtex = readGmt("gmt/gtex_signatures.gmt")
gtex.gini = readGmt("gmt/gtex_signatures_gini.gmt") 
```

### Signature Length
We compare the lengths of the signatures derived by the two methods. 
```{r compare_signature_length, fig.width=8, fig.height=3, fig.cap="**Left**: boxplot comparing the signature lengths of BioQC, and the signature sets derived from GTEx with limma and gini-index respectively. **Middle, Right**: quantile-quantile plots of the lenghts of the BioQC signatures and the signatures sets derived from GTEx with limma and gini-index respectively."}
sigLength = function(gmt) {
  return(unlist(lapply(gmt, function(x) {return(length(x$genes))})))
}
gg_qq = function(data, x, y) {
  qqr = qqplot(data[data[,1]==x,2], data[data[,1]==y,2], plot.it=FALSE)
  data.2 = data.frame(x=qqr$x, y=qqr$y)
  colnames(data.2) = c(x, y)
  coef = coef(rlm(data.2[,2] ~ data.2[,1]))
  return(ggplot(data.2, aes(x=get(x), y=get(y))) +
              geom_abline(intercept=coef[1], slope=coef[2], color="grey") +
              geom_point() + 
              labs(x=x, y=y))
}

length.bioqc = sigLength(bioqc)
length.gtex = sigLength(gtex)
length.gtex.gini = sigLength(gtex.gini)
box.data = rbind(
  data.frame(signature=rep("bioqc", length(length.bioqc)), length=length.bioqc),
  data.frame(signature=rep("gtex limma", length(length.gtex)), length=length.gtex),
  data.frame(signature=rep("gtex gini", length(length.gtex.gini)), length=length.gtex.gini))

boxp = ggplot(box.data, aes(x=signature, y=length)) + geom_boxplot()
boxp = boxp + coord_cartesian(ylim = c(0, 2500))

grid.arrange(boxp, 
             gg_qq(box.data, "bioqc", "gtex limma"), 
             gg_qq(box.data, "bioqc", "gtex gini"), ncol=3)
```
The signatures from GTEx appear to be substantially larger than the ones shipped with BioQC. Particularliy those generated with gini-index. Essentially, this boils down to a threshold problem. Adjusting the log-fold-change cutoff or the minimal required gini-index in `create_gtex_signatures.Rmd` has a direct impact on the signature sizes. The quantile-quantile plots indicate, that signatures grow linearely whith a less stringent threshold. 


### Compare the signature sets using jaccard index
```{r jaccard_index, results='hide', fig.width=8, fig.height=3, fig.cap="Heatmaps showing the jaccard index between the different signatures. Open the PDF files to see the figure with labels: [GTEx limma:GTEx gini](compare_signatures_files/jaccard_gtex-limma_gtex-gini.pdf), [GTEx limma:BioQC](compare_signatures_files/jaccard_gtex-limma_bioqc.pdf), [GTEx gini:BioQC](compare_signatures_files/jaccard_gtex-gini_bioqc.pdf)"}
pairwiseJaccMat = function(gmt1, gmt2, name1, name2) { 
  jacc.mat = matrix(nrow=length(gmt1),
                  ncol=length(gmt2),
                  dimnames=list(names(gmt1), names(gmt2)))

  for(i in 1:ncol(jacc.mat)) {
    for(j in 1:nrow(jacc.mat)) {
      jacc.mat[j,i] = jaccInd(gmt1, gmt2, names(gmt1)[j], names(gmt2)[i])
    }
  }

  jm = melt(jacc.mat)
  colnames(jm) = c(name1, name2, "jaccard_index")

  # cluster signatures
  ord1 = hclust(dist(jacc.mat))$ord
  ord2 = hclust(dist(t(jacc.mat)))$ord
  jm[,1] = factor(jm[,1], levels = rownames(jacc.mat)[ord1])
  jm[,2] = factor(jm[,2], levels = colnames(jacc.mat)[ord2])
  return(jm)
}


make.plot = function(jm, var="jaccard_index") {
  x = colnames(jm)[1]
  y = colnames(jm)[2]
  hm.palette <- colorRampPalette(rev(brewer.pal(11, 'Spectral')), space='Lab')  
  return(ggplot(jm, aes(x=get(x), y=get(y))) + 
    geom_tile(aes(fill=get(var)))+
    scale_fill_gradientn(colours = hm.palette(100)) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
    labs(x=x, y=y))
}

# pdf("fig/jaccard-index.pdf", width=25, height=10)
# print(make.plot())
# dev.off()
# 
# # make heatmap. 
# print(make.plot() + theme(axis.text=element_text(size=6)))

glgg_jm=pairwiseJaccMat(gtex, gtex.gini, "GTEx limma", "GTEx gini")
glgg = make.plot(glgg_jm)
glbqc_jm=pairwiseJaccMat(gtex, bioqc, "GTEx limma", "BioQC")
glbqc = make.plot(glbqc_jm)
gggbc_jm=pairwiseJaccMat(gtex.gini, bioqc, "GTEx gini", "BioQC")
gggbc = make.plot(gggbc_jm)

pdf("compare_signatures_files/jaccard_gtex-limma_gtex-gini.pdf", width=10, height=12)
print(glgg)
dev.off()

pdf("compare_signatures_files/jaccard_gtex-limma_bioqc.pdf", width=10, height=35)
print(glbqc)
dev.off()

pdf("compare_signatures_files/jaccard_gtex-gini_bioqc.pdf", width=15, height=35)
print(gggbc)
dev.off()

th = theme(axis.ticks.x=element_blank(),
           axis.ticks.y=element_blank(), 
           axis.text.x=element_blank(),
           axis.text.y=element_blank(), legend.position="none")

grid.arrange(glgg + th, glbqc + th, gggbc + th,ncol=3)
```

```{r pvalue, include=FALSE}
### (not used)
pval.raw = 1e-20
pval = pval.raw / (length(gtex) * length(bioqc))
pairwiseFisherMat = function(gmt1, gmt2, name1, name2) {
  fisher.mat = matrix(nrow=length(gmt1),
                  ncol=length(gmt2),
                  dimnames=list(names(gmt1), names(gmt2)))

  for(i in 1:ncol(fisher.mat)) {
    for(j in 1:nrow(fisher.mat)) {
      fisher.mat[j, i] = fisherTest(gmt1, gmt2, names(gmt1)[j], names(gmt2)[i])
    }
  }
  fm = melt(fisher.mat)
  colnames(fm) = c(name1, name2, "pvalue")
  return(fm)
}
```

Here we filter the matrix for having at least one pair of signatures per row having a jaccard index of `r jacc.thres = .2`
```{r filterd_jacc_mat}
bqc.sig = unique(data.table(glbqc_jm)[jaccard_index > jacc.thres,BioQC])
glpqc_jm.sig = glbqc_jm[glbqc_jm$BioQC %in% bqc.sig,]
make.plot(glpqc_jm.sig)
```


## R Session Info
```{r session_info}
sessionInfo()
```

