---
title: Deriving tissue signatures from the GTEx dataset. 
output:
  html_document: default 
---

The *GTEx-project*[^1] is a data resource providing high quality gene expression data from multiple reference human tissues. In this document, we identify differentially expressed genes for each tissue using [*voom*](http://bioconductor.org/packages/release/bioc/html/limma.html) and 
[*limma*](http://bioconductor.org/packages/release/bioc/html/limma.html) to build tissue-specific signatures. 

These signatures can be used with [*BioQC*](https://accio.github.io/BioQC/) to detect tissue heterogeneity in gene expression data. 

[^1]: Lonsdale, J., Thomas, J., Salvatore, M., Phillips, R., Lo, E., Shad, S., … Moore, H. F. (2013). The Genotype-Tissue Expression (GTEx) project. Nature Genetics, 45(6), 580–585. http://doi.org/10.1038/ng.2653

```{r load_libs, include=FALSE}
library(ribiosUtils)
libordie(readr)
libordie(ribiosUtils)
libordie(ribiosIO)
libordie(biomaRt)
libordie(edgeR)
libordie(data.table)
libordie(stringr)
options(fig_caption=TRUE)
library(knitr)
opts_chunk$set(out.extra='style="display:block; margin: auto"', fig.align="center") 
```

### Load the data
We read the raw counts from the `.gct` file and preprocessed tissue annotations from `GTEx-UDISDataSetID5681-sampleAnnotation.txt`. 
```{r read_data}
gct = read_gct_matrix("GTEX_v6/GTEx_Analysis_v6_RNA-seq_RNA-SeQCv1.1.8_gene_reads.gct")
annotation = read_tsv("data/GTEx-UDISDataSetID5681-sampleAnnotation.txt")
annotation.tissue = read_tsv("GTEX_v6/v6_UDIS.txt")
ids = colnames(gct)
annotation.matched = matchColumn(ids, annotation, "Experiment name")
annotation.tissue.matched = matchColumn(ids, annotation.tissue, "SAMPID")

colData = cbind(annotation.matched, annotation.tissue.matched)
rownames(colData) = colData[["Experiment name"]]

#filter rows with NA
# RIN = RNA quality index
# SMTS = tissue
validInds = which(!apply(is.na(colData[,c("RIN", "Gender", "SMTS")]), 1, sum) >= 1)
gct = gct[,validInds]
colData = colData[validInds,]
```

### perform differential analysis
We run a differntial expression analysis for tissue on the GTEx dataset with 
* Gender 
* RIN (RNA quality index)
as covariates. 

We filter that data for having at least 10 samples per gene having a count-per-milltion (CPM) of at least 1. 

We call a differetially expressed if it meets both of the following criteria: 
* fold-change of >100 (log2: `r round(log2(100), 2)`)
* p-value < 0.01 (Bonferroni corrected)
```{r limma_model}
# read file from cache if available. 
if(file.exists("limma_model.Rdata")) {
  load("limma_model.Rdata")
} else {
  # RIN = RNA quality index
  # SMTS = tissue
  dgList = DGEList(counts=gct, genes=rownames(gct))
  designMat = model.matrix(~colData$SMTS + colData$RIN + colData$Gender)
  
  # filter for cpm > 1 for at least 10 samples per gene
  CPM = cpm(dgList)
  cpmCnt = apply(CPM > 1, 1, sum)
  cpmInds = cpmCnt >= 10
  dgList.f = dgList[cpmInds,]
  
  dgList.f = calcNormFactors(dgList.f)
  
  v = voom(dgList.f, designMat, plot=FALSE)
  
  fit = lmFit(v, designMat)
  save(fit, file="limma_model.Rdata")
}

# apply cutoffs. 
fit.e = treat(fit, lfc=log2(100))
rslt = decideTests(fit.e, p.value=(0.01 / nrow(gct)))
```


### Annotate Gene Symbols
As the *BioQC* signatures use gene symbols, we need to convert the ENSEMBL gene ids to HGNC gene symbols. 
We achieve this using the [*biomaRt*](https://bioconductor.org/packages/release/bioc/html/biomaRt.html) Bioconductor package. 

```{r annotation}
ensembl = useMart("ENSEMBL_MART_ENSEMBL", dataset="hsapiens_gene_ensembl", host="www.ensembl.org")  
ensemble2hgnc = data.table(getBM(mart=ensembl, attributes=(c("ensembl_gene_id", "hgnc_symbol"))))
ensemble2hgnc = ensemble2hgnc[hgnc_symbol!='',]

# remove the .digit from the gene name. 
salitizeEnid = function(x) { 
  return(str_split(x, "\\.")[[1]][1])
}

# append sanitized EnsemblID to data table 
rslt.tab = data.table(rslt, keep.rownames=TRUE)
names = make.names(colnames(rslt.tab))
setnames(rslt.tab, names)
rn.2 = lapply(rslt.tab[,rn], salitizeEnid)
rslt.tab = rslt.tab[,rn2:=rn.2]

# append hgnc to data table 
rslt.matched = data.table(matchColumn(ensemble2hgnc$ensembl_gene_id, rslt.tab, 'rn2'))
rslt.matched = rslt.matched[,hgnc:=ensemble2hgnc[,hgnc_symbol]]
```

### Create signatures
Finally, we create lists of signatures for each tissue by collating the list of gene symbols that
are marked as overexpressed by *limma*. 
```{r create_signatures}
signatures = list()
sigCols =  names[grepl("^colData", names)]
sigCols = sigCols[1:(length(sigCols)-2)] # remove RIN and Gender
for(tCol in sigCols) {
  signatures = append(signatures, list(rslt.matched[get(tCol)==1,hgnc]))
}

names(signatures) = str_match(sigCols, "colData.SMTS(.*)")[,2]

save(signatures, file="gtex_signatures.Rdata")
```

## R Session Info
```{r session_info}
sessionInfo()
```

